---
title: "IODS-final"
author: "Mohamadali Ahi; mohamadali.ahi@helsinki.fi"
date: "15 joulukuuta 2017"
output:
  html_document:
    df_print: paged

---

In the assignment the follwing are going to be conducted: 

# hypotheses
The hypotheses are as follows:
H1. The quality of family relationship and the amount of free time after school positively affect students' final grade.
H2. Final grade can be classified by the time spent studying; as such, the higher the grade, the higher the study time.  

We first start with the first hypothesis. For that, we need to read the Student Performance dataset. I have already created a dataset called 'fam', which is a cobmination of observation from students in two classes: Math and Portuguese. Before that, however, let us access the libraries we need.

```{r}
library(tidyr)

library(dplyr)

library(ggplot2)

library(GGally)
```

Now I read the data into R.

```{r}
fam <- read.table("C:/Data/IODS-final/fam.csv", sep=",", header=TRUE)
```

The first hypothesis can be tested using a linear regression. This is the first part of my assignment. 

# 1. Multiple Linear Regression 

To know the dataset well, I first look at the names of the variables.

```{r}
colnames(fam)
```

Before proceeding to the regression, it would be a good idea to look at the relationship between variables and their distributions.   

```{r}
p <- ggpairs(fam, mapping = aes (col = sex, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

p
```

```{r}
pairs(fam)
```
